Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-09 paper/paper.tex
Rewording proof of adequacy and normalization

\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}

@@

\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
  \label{lemma:parantirenaming}
  If $[[a < xi > => b0]]$, then there exists some $[[b]]$ such that
  $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

@@

Before we can prove the fundamental theorem and derive the normalization
property as its corollary, we need to additionally formulate and prove an
\emph{adequacy property} about the logical relation.  This property, that the
interpretation of each type is a \emph{reducibility candidate}, allows us to
conclude that every term in each interpretation has a normal form. In the
previous section, we only needed a property of the interpretation of the
$[[Void]]$ type. However, for this section, we need to know something about
the interpretation of every type.

Furthermore, to prove this adequacy property, we need to strengthen it to also
give us more information about neutral terms as we proceed by induction. In
particular, we need to know that all terms that reduce to neutral forms are
contained within the interpretation.
Therefore, we formally define when a set is a \emph{reducibility candidate}
(shortened as $CR$) as follows. Our definition of $CR$ is inspired by
\citet{girard1989proofs}, but not identical since we only care about
weak normalization.
\begin{definition}[Reducibility Candidates (CR)\footnote{\dotv{semtypingopen.v}{CR}}]
  Let $[[S]]$ be a set of terms. We say that $[[S]] \in CR$
  if and only if conditions $CR_1$ and $CR_2$ hold.
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff\ [[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
  \end{itemize}
\end{definition}

We now state and prove the adequacy lemma.
\begin{lemma}[Adequacy\footnote{\dotv{semtypingopen.v}{adequacy}}]
  \label{lemma:adequacy}
  If $[[InterpR i A S]]$, then we have $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  We start by strong induction over $[[i]]$. We are given the
  induction hypothesis that for all $[[j < i]]$, $[[InterpR j A S]]$
  implies $[[S]] \in CR$. Our goal is to show $[[InterpR i A S]]$
  implies $[[S]] \in CR$.

  By Definition~\ref{fig:logrelrec}, we have the
  equality $[[InterpR i A S]] = [[Interp I i A S]]$  where $[[I i]] :=
  [[{A | exists S , InterpR i A S}]]$.
  We then proceed by structural induction over the derivation of $[[Interp I i A
  S]]$. The only interesting cases are \rref{I-Pi} and \rref{I-Set}.
  The function case requires Lemmas~\ref{lemma:extwn} and
  \ref{lemma:wnewn}, which we have shown earlier.

  The \rref{I-Set} case is the most interesting. We must show that
  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
  has the same definition from earlier but its domain restricted to
  numbers less than $[[j]]$. We perform another induction on
  the derivation of $[[Interp I j A S]]$. All cases are trivial except for the case for
  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
  A]]$. To derive $[[wn (Pi x : A . B)]]$, it remains to show $[[wn B]]$. We
  use the outermost induction hypothesis to show that an arbitrary $[[y]]$
  semantically inhabits $[[A]]$, from which we derive $[[wn (B {y / x})]]$
  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
\end{proof}

The formulation of semantic well-typedness and the fundamental lemma from
Section~\ref{sec:logrelproof} remains
unchanged.
The proof of the fundamental lemma\footnote{\dotv{soundnessopen.v}{soundness}} is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemma~\ref{lemma:adequacy}, the adequacy property.

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, which injects
  variables as terms. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
  conclude that $[[wn a]]$ and $[[wn A]]$.
\end{proof}

>>>>

\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
\newcommand{\CR}{\mathit{CR}}

@@

\begin{lemma}[Par antirenaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
  \label{lemma:parantirenaming}
  If $[[a < xi > => b0]]$, then there exists some $[[b]]$ such that
  $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

@@

Before we can prove the fundamental theorem and derive the normalization
property as its corollary, we need to formulate and prove an \emph{adequacy}
property of the logical relation, which states that the interpretation of each
type is a \emph{reducibility candidate} ($\CR$), which characterizes sets of
weakly normalizing terms. In contrast to proving consistency, which only
requires knowing that the interpretation of $[[Void]]$ is empty, adequacy
allows us to conclude that all terms in \emph{every} interpretation have a
normal form.

To prove adequacy, we need to strengthen it to with more information about
neutral terms as we proceed by induction. In particular, we need to know that
all terms that reduce to neutral forms are contained within the interpretation.
Therefore, we define reducibility candidates as follows, inspired by
\citet{girard1989proofs}, but modified for weak normalization only. Adequacy
then states that interpretations are in $\CR$.

\begin{definition}[Reducibility Candidates ($\CR$)\footnote{\dotv{semtypingopen.v}{CR}}]
  A set of terms $[[S]]$ is in $\CR$
  if and only if conditions $\CR_1$ and $\CR_2$ hold.
  \begin{itemize}
  \item $[[S]] \in \CR_1 \iff\ [[forall a, (# wne a implies a in S #)]]$
  \item $[[S]] \in \CR_2 \iff\ [[forall a, (# a in S implies wn a #)]]$
  \end{itemize}
\end{definition}

\begin{lemma}[Adequacy\footnote{\dotv{semtypingopen.v}{adequacy}}]
  \label{lemma:adequacy}
  If $[[InterpR i A S]]$, then $[[S]] \in \CR$.
\end{lemma}

\begin{proof}
  We start by strong induction over $[[i]]$. We are given the
  induction hypothesis that for all $[[j < i]]$, $[[InterpR j A S]]$
  implies $[[S]] \in \CR$. Our goal is to show that $[[InterpR i A S]]$
  implies $[[S]] \in \CR$.

  By Definition~\ref{fig:logrelrec}, we have $[[Interp I i A S]]$,
  where $[[I j]] := [[{A | exists S , InterpR j A S}]]$ for $[[j < i]]$.
  We then proceed by structural induction on the derivation of
  $[[Interp I i A S]]$. The interesting cases are \rref{I-Pi} and \rref{I-Set}.
  For \rref{I-Pi}, proving $S \in \CR_1$ requires Lemma~\ref{lemma:wnewn}. When
  proving $S \in \CR_2$, by the structural induction hypotheses, we have that
  $[[wn (b a)]]$ for any $[[wne a]]$, while the goal is to show that $[[wn b]]$.
  We pick an arbitrary $[[x]]$ and conclude by applying Lemma~\ref{lemma:extwn}
  to $[[b x]]$.

  The \rref{I-Set} case is the most interesting. We must show that
  for all $[[j < i]]$, $[[{A | exists S, InterpR j A S}]] \in \CR$.
  We immediately know that $[[{A | exists S, InterpR j A S}]] \in \CR_1$
  by \rref{I-Ne}. It remains to show that
  $[[{A | exists S, InterpR j A S}]] \in \CR_2$, or equivalently that
  $[[Interp I j A S]]$ implies $[[wn A]]$, where $[[I]]$ is defined as above
  over $k < j$.

  We assume $[[Interp I j A S]]$ and proceed once more by structural induction
  on its derivation. All cases are trivial except for \rref{I-Pi}. The induction
  hypothesis immediately gives $[[wn A]]$. To derive $[[wn (Pi x : A . B)]]$,
  it remains to show $[[wn B]]$. We use the outermost induction hypothesis on
  $[[Interp I j A S]]$ to show that an arbitrary $[[y]]$ semantically inhabits
  $[[A]]$ by $\CR_1$, followed by the structural induction hypothesis to derive
  $[[wn (B {y / x})]]$. We conclude $[[wn B]]$ by applying antirenaming
  (Lemma~\ref{lemma:parantirenaming}).
\end{proof}

The formulation of semantic well-typedness and the fundamental lemma from
Section~\ref{sec:logrelproof} remains unchanged. The proof of the fundamental
lemma\footnote{\dotv{soundnessopen.v}{soundness}} is still carried out by
induction over the typing derivation, where the additional neutral term related
cases are handled by adequacy. The normalization property then follows as a
corollary of the fundamental theorem.

\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}

\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, and the
  condition $[[idtm |= G]]$ is satisfied by adequacy, which says that variables,
  as neutral terms, semantically inhabit the interpretations of the types in
  $[[G]]$. Performing the identity substitutions, we then know that there exist
  $[[i]]$ and $[[S]]$ such that $[[InterpR i A S]]$ and $[[a in S]]$. Finally,
  by adequacy again, we conclude that $[[wn a]]$ and $[[wn A]]$.
\end{proof}
